---
title: "Group Task"
author: "Mencía Gómez Luna (DNI: 05468738-M), Laura Toro Iglesias (DNI: 06591204-W), Alba García Vega (DNI: 71904192M), Julia Martyniewicz(DNI: CAJ641749)"
format:
  html:
    theme: [style.scss]
    toc: true
    toc-location: right
    toc-title: Índice
    embed-resources: true
---

## Instructions (read before starting)

-   Modify within the `.qmd` document your personal data (names and ID) located in the header of the file.

-   Make sure, **BEFORE further editing** the document, that the `.qmd` file is rendered correctly and the corresponding `.html` is generated in your local folder on your computer.

-   The chunks (code boxes) created are either empty or incomplete. Once you edit what you consider, you must change each chunk to `#| eval: true` (or remove it directly) for them to run.

-   Remember that you can run chunk by chunk with the *play* button or run all chunks up to a given chunk (with the button to the left of the previous one).

### Required packages

> Insert in the lower chunk the packages you will need

```{r}
rm(list = ls())
library(tidyverse)
```

## Group task

### Data

The practice will be based on the **electoral data archives** below, compiling data on elections to the Spanish Congress of Deputies from 2008 to the present, as well as surveys, municipalities codes and abbreviations.

```{r}
# NO TOQUES NADA
election_data <- read_csv(file = "./data/datos_elecciones_brutos.csv")
cod_mun <- read_csv(file = "./data/cod_mun.csv")
surveys <- read_csv(file = "./data/historical_surveys.csv")
abbrev <- read_csv(file = "./data/siglas.csv")
```

The data will be as follows:

-   `election_data`: file with election data for Congress from 2018 to the last ones in 2019.

    -   `tipo_eleccion`: type of election (02 if congressional election)
    -   `anno`, `mes`: year and month of elections
    -   `vuelta`: electoral round (1 if first round)
    -   `codigo_ccaa, codigo_provincia, codigo_municipio, codigo_distrito_electoral`: code of the ccaa, province, municipality and electoral district.
    -   `numero_mesas`: number of polling stations
    -   `censo`: census
    -   `participacion_1, participacion_2`: participation in the first preview (14:00) and second preview (18:00) before polls close (20:00)
    -   `votos_blancos`: blank ballots
    -   `votos_candidaturas`: party ballots
    -   `votos_nulos`: null ballots
    -   ballots for each party

-   `cod_mun`: file with the codes and names of each municipality

-   `abbrev`: acronyms and names associated with each party

-   `surveys`: table of electoral polls since 1982. Some of the variables are the following:

    -   `type_survey`: type of survey (national, regional, etc.)
    -   `date_elec`: date of future elections
    -   `id_pollster`, `pollster`, `media`: id and name of the polling company, as well as the media that commissioned it.
    -   `field_date_from`, `field_date_to`: start and end date of fieldwork
    -   `exit_poll`: whether it is an exit poll or not
    -   `size`: sample size
    -   `turnout`: turnout estimate
    -   estimated voting intentions for the main parties

### Objectives and mandatory items

The objective of the delivery is to perform an analysis of the electoral data, carrying out the debugging, summaries and graphs you consider, both of their results and the accuracy of the electoral polls.

Specifically, **you must work only in the time window that includes the elections from 2008 to the last elections of 2019**.

#### General comments

In addition to what you see fit to execute, the following items are mandatory:

-   Each group should **present before 9th January (23:59) an analysis of the data** in `.qmd` and `.html` format in **Quarto slides** mode, which **will be the ones they will present on the day of the presentation**.

- **Quarto slides should be uploaded to Github** (the link should be provided by a member of each group).

- The **maximum number of slides** should be 40. The **maximum time for each group** will be 20-22 minutes (+5 minutes for questions).

-   During the presentation you will **explain (summarised!) the analysis performed** so that **each team member speaks for a similar amount of time** and **each member can be asked about any of the steps**. The grade does not have to be the same for all members. 

-   It will be valued not only the content but also the container (aesthetics).

-   The **objective is to demonstrate that the maximum knowledge of the course has been acquired**: the more content of the syllabus is included, the better.

#### Mandatory items:

1. Data should be **converted to tidydata** where appropriate.

2. You should **include at least one join** between tables.

3. Reminder: information = variance, so **remove columns that are not going to contribute anything**.

4. The **glue and lubridate** packages should be used at some point, as well as the **forcats**. The use of **ggplot2** will be highly valued.

5. The following should be used at least once:
    -   mutate
    -   summarise
    -   group_by (or equivalent)
    -   case_when

6. We have many, many parties running for election. **We will only be interested in the following parties**:

    -   PARTIDO SOCIALISTA OBRERO ESPAÑOL (beware: it has/had federations - branches - with some other name).
    -   PARTIDO POPULAR
    -   CIUDADANOS (caution: has/had federations - branches - with some other name)
    -   PARTIDO NACIONALISTA VASCO
    -   BLOQUE NACIONALISTA GALLEGO
    -   CONVERGÈNCIA I UNIÓ
    -   UNIDAS PODEMOS - IU (beware that here they have had various names - IU, podem, ezker batua, ...- and have not always gone together, but here we will analyze them together)
    -   ESQUERRA REPUBLICANA DE CATALUNYA
    -   EH - BILDU (are now a coalition of parties formed by Sortu, Eusko Alkartasuna, Aralar, Alternatiba)
    -   MÁS PAÍS
    -   VOX

7. Anything other than any of the above parties should be imputed as "OTHER". Remember to add properly the data after the previous recoding.

8.   Party acronyms will be used for the visualizations. The inclusion of graphics will be highly valued (see <https://r-graph-gallery.com/>).

9. You must use all 4 data files at some point.

10.  You must define at least one (non-trivial) function of your own.

11.  You will have to discard mandatory polls that:

    -   refer to elections before 2018
    -   that are exit polls
    -   have a sample size of less than 750 or are unknown
    -   that have less than 1 or less fieldwork days

12.   You must obligatorily answer the following questions (plus those that you consider analyzing to distinguish yourself from the rest of the teams, either numerically and/or graphically)

    -   Which party was the winner in the municipalities with more than 100,000 habitants (census) in each of the elections?
    -   Which party was the second when the first was the PSOE? And when the first was the PP?
    -   Who benefits from low turnout?
    -   How to analyze the relationship between census and vote? Is it true that certain parties win in rural areas?
    -   How to calibrate the error of the polls (remember that the polls are voting intentions at national level)?
    -   Which polling houses got it right the most and which ones deviated the most from the results?

**You should include at least 3 more "original" questions** that you think that it could be interesting to be answer with the data.

#### Marks

**The one who does the most things will not be valued the most**. More is not always better. The **originality** (with respect to the rest of the works, for example in the analyzed or in the subject or ...) of what has been proposed, in the handling of tables (or in the visualization), the **caring** put in the delivery (care in life is important) and the **relevance** of what has been done will be valued. Once you have the mandatory items with your database more or less completed, **think before** chopping code: what could be interesting? What do I need to get a summary both numerical and visual?

Remember that the real goal is to demonstrate a mastery of the tools seen throughout the course. And that happens not only by the quantity of them used but also by the quality when executing them.

**Some dataviz will be extremely positive valued**.


## DATA PREPARATION

First, we **convert the data to tidy data**: In both "election_data" and "surveys" databases, each party appears as a variable; so we create a new variable that unites all these parties so they appear as categories of a single variable "party", and the values appear in a new variable called "estimated_voting" and "ballots" respectively.

```{r}
#| code-fold: true
surveys_tidy <- surveys |>
  pivot_longer(
    cols = 11:59,
    names_to = "party",      
    values_to = "estimated_voting"    
  )

election_data_tidy <- election_data |>
  pivot_longer(
    cols = 16:471,
    names_to = "party",      
    values_to = "ballots"
  )

```

Now we eliminate the non relevant parties (according to points 6 and 7) for "surveys_tidy" and "election_data_tidy"

```{r}
#| code-fold: true
library(dplyr)
important_parties <- c(
  "PSOE", "PP", "CS", "BNG", "EAJ-PNV", "CIU", 
  "PODEMOS", "IU", "UP", "EH-BILDU", "ERC", "VOX", "EE", "EA",
  "EH", "PCE", "MP"
)

surveys_tidy1 <- surveys_tidy |>
  mutate(party = case_when(
    party %in% important_parties ~ party,  #Keep important parties
    TRUE ~ "Other"    #New category: other parties
  )) |>
  
  #Change some party names
  mutate(party = case_when(
    party == "PODEMOS" ~ "UNIDAS PODEMOS-IU",
    party == "IU" ~ "UNIDAS PODEMOS-IU",
    party == "UP" ~ "UNIDAS PODEMOS-IU",
    party == "PCE" ~ "PODEMOS-IU",
    party == "MUC" ~ "PODEMOS-IU", #partido integrado en IU, aunque muy minoritario
    party == "ICV" ~ "PODEMOS-IU", #verdes en valencia, van con podemos
    party == "EE" ~ "PSOE", #PSOE in País Vasco
    party == "EA" ~ "EH-BILDU",
    party == "EH" ~ "EH-BILDU",
    party == "COMPROMIS" ~ "PODEMOS-IU",
    party == "JXCAT-JUNTS" ~ "CIU",
    party == "PDECAT" ~ "CIU",
    TRUE ~ party  #Keep the rest without changes
  ))
```

We check which parties we have eliminated from the first surveys_tidy 

```{r}
#| code-fold: true

setdiff(surveys_tidy$party, surveys_tidy1$party)

unique(surveys_tidy1$party)
```

We do the same for "election_data_tidy"

```{r}
#| code-fold: true
important_parties2 <- c("PARTIDO SOCIALISTA OBRERO ESPAÑOL", "PARTIDO POPULAR", "PODEMOS", "UNIDAS PODEMOS", "VOX", "ARALAR", "EUSKO ALKARTASUNA", "MÁS PAÍS", "EUSKAL HERRIA BILDU (EH Bildu)", "EN MAREA", "EQUO", "IZQUIERDA REPUBLICANA", "CONVERGÈNCIA i UNIÓ", "ESQUERRA REPUBLICANA DE CATALUNYA", "EUSKAL HERRIA BILDU", "CIUDADANOS PARTIDO DE LA CIUDADANÍA", "BLOQUE NACIONALISTA GALEGO") 

election_data_tidy1 <- election_data_tidy |>
  mutate(party = case_when(
    party %in% important_parties2 ~ party,            
    grepl("PODEMOS|UNIDAS PODEMOS|IU|IZQUIERDA UNIDA", party, ignore.case = TRUE) ~ "PODEMOS-IU", 
    grepl("PP|PARTIDO POPULAR|PARTIDO POPULAR/PARTIT POPULAR", party, ignore.case = TRUE) ~ "PARTIDO POPULAR",
    grepl("PARTIDO SOCIALISTA|PSOE", party, ignore.case = TRUE) ~ "PARTIDO SOCIALISTA OBRERO ESPAÑOL",
    grepl ("PARTIDO NACIONALISTA VASCO", party, ignore.case = TRUE) ~ "PARTIDO NACIONALISTA VASCO",
    grepl ("PARTIDO DE LA CIUDADANÍA", party, ignore.case = TRUE) ~ "CIUDADANOS PARTIDO DE LA CIUDADANÍA", #Important: to include all the "Ciudadanos" parties we don't use the word "Ciudadanos" because this would include the wrong parties: Ciudadanos en blanco, Ciudadanos rurales agrupados, etc
    TRUE ~ "Other"       
  ))


election_data_tidy2 <- election_data_tidy1|> #Change some party names
  mutate(party = case_when(
    party == "PODEMOS" ~ "PODEMOS-IU",
    party == "EN MAREA" ~ "PODEMOS-IU",
    party == "UNIDAS PODEMOS" ~ "PODEMOS-IU",
    party == "EQUO" ~ "PODEMOS-IU",
    party == "IZQUIERDA REPUBLICANA" ~ "PODEMOS-IU",
    party == "EUSKAL HERRIA BILDU (EH Bildu)" ~ "EH BILDU",
    party == "EUSKAL HERRIA BILDU" ~ "EH BILDU",
    party == "ARALAR" ~ "EH BILDU",
    party == "EUSKO ALKARTASUNA" ~ "EH BILDU",
    TRUE ~ party  #Keep the rest without changes
  ))

unique(election_data_tidy2$party) #To check the results 
```


Now, according to point 11, we have to discard some polls, so we filter:

```{r}
#| code-fold: true
library(lubridate)
surveys_tidy2 <- surveys_tidy1 |>
  mutate(
    date_elec = ymd(date_elec),
    field_date_from = ymd(field_date_from),
    field_date_to = ymd(field_date_to),
    fieldwork_days = as.numeric(difftime(field_date_to, field_date_from, units = "days")) # We calculate the fieldwork days
  ) |>
  filter(
    date_elec >= "2008-01-01",    # Elections >= 2008
    exit_poll == "FALSE",         # No exit polls
    size >= 750,                 # Size >= 750  and known
    fieldwork_days > 1            # Fieldwork days > 1 
    )
```


JOINS:

Possible joins that may be useful:

Joining cod_mun and election_data_tidy2 to have the full name of municipalities (will be useful for question 1). First we create a new variable in election_data_tidy2 that is the full code of the municipality, using the columns "codigo_ccaa", "codigo_provincia" and "codigo_municipio"

```{r}
#| code-fold: true
election_data_tidy_new <- election_data_tidy2 |> 
  mutate(cod_mun_full = paste(codigo_ccaa, codigo_provincia, codigo_municipio, sep = "-"))

#left join to have the names of municipalities
election_data_tidy_new <- election_data_tidy_new |> 
  left_join(cod_mun, by = c("cod_mun_full" = "cod_mun"))

```

Now we have a table with the info from election_data and cod_mun.

JOIN ABBREV:

First, we have to clean the names from abbrev:

```{r}
#| code-fold: true
important_parties3 <- c("PARTIDO SOCIALISTA OBRERO ESPAÑOL", "PARTIDO POPULAR", "PODEMOS", "UNIDAS PODEMOS", "VOX", "ARALAR", "EUSKO ALKARTASUNA", "MÁS PAÍS", "EUSKAL HERRIA BILDU (EH Bildu)", "EN MAREA", "EQUO", "IZQUIERDA REPUBLICANA", "CONVERGÈNCIA i UNIÓ", "ESQUERRA REPUBLICANA DE CATALUNYA", "EUSKAL HERRIA BILDU", "CIUDADANOS-PARTIDO DE LA CIUDADANÍA", "BLOQUE NACIONALISTA GALEGO") 

abbrev_clean <- abbrev |>
  mutate(denominacion = case_when(
    denominacion %in% important_parties3 ~ denominacion,            
    grepl("PODEMOS|UNIDAS PODEMOS|IU|IZQUIERDA UNIDA", denominacion, ignore.case = TRUE) ~ "PODEMOS-IU", 
    grepl("PP|PARTIDO POPULAR|PARTIDO POPULAR/PARTIT POPULAR", denominacion, ignore.case = TRUE) ~ "PARTIDO POPULAR",
    grepl("PARTIDO SOCIALISTA|PSOE", denominacion, ignore.case = TRUE) ~ "PARTIDO SOCIALISTA OBRERO ESPAÑOL",
    grepl ("PARTIDO NACIONALISTA VASCO", denominacion, ignore.case = TRUE) ~ "PARTIDO NACIONALISTA VASCO",
    grepl ("CIUDADANOS PARTIDO DE LA CIUDADANIA", denominacion, ignore.case = TRUE) ~ "CIUDADANOS PARTIDO DE LA CIUDADANÍA",
    grepl ("CONVERGENCIA", denominacion, ignore.case = TRUE) ~ "CONVERGÈNCIA i UNIÓ",
    TRUE ~ "Other"                                   
  ))

abbrev_clean2 <- abbrev_clean |>
  mutate(denominacion = case_when(
    denominacion == "PODEMOS" ~ "PODEMOS-IU",
    denominacion == "EN MAREA" ~ "PODEMOS-IU",
    denominacion == "UNIDAS PODEMOS" ~ "PODEMOS-IU",
    denominacion == "EQUO" ~ "PODEMOS-IU",
    denominacion == "IZQUIERDA REPUBLICANA" ~ "PODEMOS-IU",
    denominacion == "EUSKAL HERRIA BILDU (EH Bildu)" ~ "EH BILDU",
    denominacion == "EUSKAL HERRIA BILDU" ~ "EH BILDU",
    denominacion == "ARALAR" ~ "EH BILDU",
    denominacion == "EUSKO ALKARTASUNA" ~ "EH BILDU",
    denominacion == "CIUDADANOS-PARTIDO DE LA CIUDADANÍA" ~ "CIUDADANOS PARTIDO DE LA CIUDADANÍA",
    TRUE ~ denominacion
  ))

unique(abbrev_clean2$denominacion) #To check the results

#Cambiar las siglas
abbrev_clean2 <- abbrev_clean2  |> 
  mutate(siglas = case_when(
    denominacion == "PARTIDO SOCIALISTA OBRERO ESPAÑOL" ~ "PSOE",
    denominacion == "PARTIDO POPULAR" ~ "PP",
    denominacion == "PODEMOS-IU" ~ "PODEMOS-IU",
    denominacion == "EH BILDU" ~ "EH-BILDU",
    denominacion == "ESQUERRA REPUBLICANA DE CATALUNYA" ~ "ERC",
    denominacion == "BLOQUE NACIONALISTA GALEGO" ~ "BNG",
    denominacion == "VOX" ~ "VOX",
    denominacion == "MÁS PAÍS" ~ "MP",
    denominacion == "PARTIDO NACIONALISTA VASCO" ~ "EAJ-PNV",
    denominacion == "CIUDADANOS PARTIDO DE LA CIUDADANÍA" ~ "CS",
    denominacion == "Other" ~ "Other", #antes estaba en NA 
    denominacion == "CONVERGÈNCIA i UNIÓ" ~ "CIU",
    TRUE ~ siglas
  ))

unique(abbrev_clean2$siglas)

#remove duplicates for the join
abbrev_clean2_unique <- abbrev_clean2 |> 
  distinct(denominacion, .keep_all = TRUE)
```

Then we do the join, which we can do both with the survey and the election data, so we have a more complete databases. We also reorder and rename some of the new variables so it's clearer:

```{r}
#| code-fold: true

#election data
final_election_data <- election_data_tidy_new |> 
  left_join(abbrev_clean2_unique, by = c("party" = "denominacion"))

final_election_data <- final_election_data |> 
  relocate(siglas, .after = party)


#survey data

final_survey_data <- surveys_tidy2 |> 
  left_join(abbrev_clean2_unique, by = c("party" = "siglas"))

final_survey_data <- final_survey_data |> 
  relocate(denominacion, .before = party) |> 
  rename(
    abbrev = party,
    party = denominacion
  )

final_survey_data <- final_survey_data |> 
  mutate(party = if_else(is.na(party), "Other", party))

```



### QUESTIONS

### Question 1

Which party was the winner in the municipalities with more than 100,000 habitants (census) in each of the elections?

Anotación: creo que la pregunta está bien resuelta, pero los resultados no tienen sentido. Podemos sale ganador en muchos sitios y no tiene sentido. No sé si el fallo viene de las tablas de election data o de después.
Alba: pues a mí me cuadra. Ten en cuenta que son municipios y que estás juntando muchos partidos. Yo creo que está bien. De todas formas se revisa, porque si hay algo más es la base de datos arreglada como tal, así que sería cambiar la base y volver a hacer run. 

```{r}
#| code-fold: true

#we filter the large municipalities
large_municipalities <- final_election_data |> 
  filter(censo > 100000)


votes_by_party <- large_municipalities |> 
  group_by(municipio, party, anno, mes) |> 
  summarise(total_votes = sum(ballots, na.rm = TRUE), .groups = "drop")

print(votes_by_party) 

winners_by_municipality <- votes_by_party |> 
  group_by(municipio, anno, mes) |> 
  filter(total_votes == max(total_votes)) |> 
  ungroup()

print(winners_by_municipality) 


```



### Question 2

Which party was the second when the first was the PSOE? And when the first was the PP?

```{r}
#| code-fold: true
library(dplyr)

votes_by_party_all <- final_election_data |> 
  group_by(municipio, siglas, anno, mes) |> 
  summarise(total_votes = sum(ballots, na.rm = TRUE), .groups = "drop")

winners_by_municipality_all <- votes_by_party_all |> 
  group_by(municipio, anno, mes) |> 
  filter(total_votes == max(total_votes)) |> 
  ungroup()

# Second when PSOE was first

psoe_winner <- winners_by_municipality_all |> filter(siglas == "PSOE")

second_place_psoe_first <- votes_by_party_all |> filter(municipio %in% psoe_winner$municipio) |> 
  filter(!(siglas == "PSOE")) |> 
  group_by(municipio, anno, mes) |> 
  filter(total_votes == max(total_votes)) |> 
  ungroup()

results_psoe_winner <- psoe_winner |>
  rename(winner_party = siglas, winner_votes = total_votes) |> 
  left_join(
    second_place_psoe_first |> 
      rename(second_party = siglas, second_votes = total_votes),
    by = c("municipio", "anno", "mes")
  ) |> 
  relocate(anno, mes, .after = municipio)

results_psoe_winner #Results in column "second_party"


```


```{r}
#| code-fold: true

# Second when PP won
pp_winner <- winners_by_municipality_all |> filter(siglas == "PP")

second_place_pp_first <- votes_by_party_all |> filter(municipio %in% pp_winner$municipio) |> 
  filter(!(siglas == "PP")) |> 
  group_by(municipio, anno, mes) |> 
  filter(total_votes == max(total_votes)) |> 
  ungroup()

results_pp_winner <- pp_winner |> rename(winner_party = siglas, winner_votes = total_votes) |> 
  left_join(
    second_place_pp_first |> 
      rename(second_party = siglas, second_votes = total_votes),
    by = c("municipio", "anno", "mes")
  ) |> 
  relocate(anno, mes, .after = municipio)

results_pp_winner #Results in column "Second party"

```


```{r}
#| code-fold: true
second_party_year_psoe_summary <- results_psoe_winner |> 
  group_by(second_party, anno) |> 
  summarise(count = n(), .groups = "drop")

ggplot(second_party_year_psoe_summary, aes(x = factor(anno), y = count, fill = second_party)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("BNG" = "pink", "CS" = "orange",
                    "CIU" = "darkred", "EH-BILDU" = "darkgreen", "ERC" = "yellow", "Other" = "grey", "EAJ-PNV" = "lightblue", "PP" = "blue", "PODEMOS-IU" = "purple", "VOX" = "green")) +
  labs(title = "Distribution of second places by year when PSOE wins",
       x = "Year",
       y = "Number of municipalities",
       fill = "Partido") +
  theme_minimal()

```

```{r}
#| code-fold: true
second_party_year_pp_summary <- results_pp_winner |> 
  group_by(second_party, anno) |> 
  summarise(count = n(), .groups = "drop")

ggplot(second_party_year_pp_summary, aes(x = factor(anno), y = count, fill = second_party)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("BNG" = "pink", "CS" = "orange",
                    "CIU" = "darkred", "EH-BILDU" = "darkgreen", "ERC" = "yellow", "Other" = "grey", "EAJ-PNV" = "lightblue", "PP" = "blue", "PODEMOS-IU" = "purple", "VOX" = "green", "PSOE" = "red")) +
  labs(title = "Distribution of second places by year when PP wins",
       x = "Year",
       y = "Number of municipalities",
       fill = "Partido") +
  theme_minimal()
```

### Question 3

Who benefits from low turnout?
First of all, what is a low turnout? It makes reference to the electoral districts where the electoral participation is low. 
Who benefits from low turnout?
We consider low turnout <0,6 due to the turnout tendency in Spain (We can not compare with other countries because in some like USA 60% is high and in other like Sweden 60% is very very very very very low). So we compare participatory trend in Spain during democracy, where the average participation rate in general elections in Spain is 72.1%, so <70% would be lower than usual and <60% would be low.  (20 Minutos, 2023)

The parties that benefit from low turnout are: CIU, PNV, Bildu, ERC, PP, VOX and Other parties.  
>REVISAR interpretación

We can also do this with a function:

```{r}
#| code-fold: true

# Vote share: proportion of votes that a party or candidate received in relation to the total number of voters

election_data_vote_share <- final_election_data |>
  group_by(codigo_municipio) |>
  mutate(vote_share = ballots / censo)

turnout_election_data <- election_data_vote_share|>
  mutate(turnout = (participacion_1 + participacion_2) / censo) |> 
  mutate(turnout_bin = ifelse(turnout < 0.6, "Low Turnout", "High Turnout"))

analyze_low_turnout <- function(data, turnout_col = "turnout", ballots_col = "ballots", censo_col = "censo", party_col = "siglas") {
  # Calculate vote share
  data <- data |>
    group_by(codigo_municipio) |>
    mutate(vote_share = !!sym(ballots_col) / !!sym(censo_col))
  
  # Calculate turnout and bin it into low and high categories
  data <- data |>
    mutate(turnout_bin = ifelse(!!sym(turnout_col) < 0.6, 
                                "Low Turnout", 
                                "High Turnout"))
  
  # Summarize vote share by turnout category and party
  summary <- data |>
    group_by(!!sym(party_col), turnout_bin) |>
    summarise(avg_vote_share = mean(vote_share, na.rm = TRUE), .groups = "drop") |>
    arrange(desc(avg_vote_share))
  
  return(summary)
}

turnout_analysis <- analyze_low_turnout(turnout_election_data, 
                                        turnout_col = "turnout", 
                                        ballots_col = "ballots", 
                                        censo_col = "censo", 
                                        party_col = "siglas")
print(turnout_analysis)


```

Visualization:

```{r}
#| code-fold: true

ggplot(turnout_analysis, aes(x = siglas, y = avg_vote_share, fill = turnout_bin)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Vote Share by Turnout Level and Party",
       x = "Party",
       y = "Average Vote Share (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

### Question 4

How to analyze the relationship between census and vote? Is it true that certain parties win in rural areas?

In Spain is considered "rural"a population of <30000 inhabitants according to Ley 45/2007, de 13 de diciembre, «BOE» núm. 299, de 14 de diciembre de 2007, páginas 51339 a 51349 (11 págs.)

```{r}
#| code-fold: true

election_data_areas <- final_election_data |> 
  mutate(area_type = case_when(
    censo < 30000 ~ "Rural",
    TRUE ~ "Urban"
  ))  |> 
  group_by(area_type, party, siglas)  |> 
  summarise(total_votes = sum(ballots, na.rm = TRUE))  |> 
  mutate(vote_share = total_votes / sum(total_votes))

party_colors <- c(
  "BLOQUE NACIONALISTA GALEGO" = "skyblue", 
  "CIUDADANOS PARTIDO DE LA CIUDADANÍA" = "#EB6109",
  "CONVERGÈNCIA i UNIÓ" = "#F7D117",
  "EH BILDU" = "lightgreen",
  "ESQUERRA REPUBLICANA DE CATALUNYA" = "#FFB000",
  "MÁS PAÍS" = "#00B1E0",
  "Other" = "#D3D3D3",  # Color genérico para "Other"
  "PARTIDO NACIONALISTA VASCO" = "#228B22",
  "PARTIDO POPULAR" = "#1D84CE",
  "PARTIDO SOCIALISTA OBRERO ESPAÑOL" = "#FF0000",
  "PODEMOS-IU" = "#6A097D",
  "VOX" = "#5BC035"
)


```

We don't see very drastic differences, but we see how, for example, rural areas report less votes for newer parties (like Vox or Podemos, especially the latter) and more for traditional parties like PP or PSOE.
REVISAR PORQUE NO ME COGE SIGLAS

```{r}
#| code-fold: true

vote_share_table <- election_data_areas |> 
  group_by(party) |> 
  mutate(total_votes_party = sum(total_votes)) |> 
  ungroup() |> 
  mutate(vote_share_adjusted = total_votes / total_votes_party) |> 
  select(party, area_type, vote_share_adjusted) |> 
  pivot_wider(names_from = area_type, values_from = vote_share_adjusted) |> 
  mutate(Rural = Rural * 100, Urban = Urban * 100) |> 
  select(party, Rural, Urban)

vote_share_table
```

```{r}
#| code-fold: true

# Convertir la tabla a formato largo para ggplot 
vote_share_long <- vote_share_table |>  
  pivot_longer(cols = c(Rural, Urban), names_to = "area_type", values_to = "vote_share")

# Crear el gráfico de barras apiladas
ggplot(vote_share_long, aes(x = party, y = vote_share, fill = area_type)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = c("Rural" = "#8FCB9B", "Urban" = "#4A90E2")) +
  labs(title = "Vote Share by Party and Area Type (100%)", 
       x = "Party", 
       y = "Vote Share (%)", 
       fill = "Area Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Question 5

How to calibrate the error of the polls (remember that the polls are voting intentions at national level)?

```{r}
#| code-fold: true
library(stringr)

# First, let's compute the total votes for each party in each election
total_votes_by_party <- final_election_data |>
  group_by(anno, mes, party) |>
  summarize(total_ballots = sum(ballots, na.rm = TRUE), .groups = "drop") |> 
   group_by(anno, mes) |> 
   mutate(total_ballots_election = sum(total_ballots, na.rm = TRUE)) |> 
    ungroup() |> 
   mutate(real_percentage = total_ballots / total_ballots_election) |> 
  select(anno, mes, party, real_percentage) |> 
  mutate(mes = as.integer(mes)) # Convert mes to integer (there was a problem without it before)

# Now, let's average survey data for each party for each election
survey_estimates <- final_survey_data |>
  filter(!is.na(estimated_voting)) |> 
  group_by(date_elec, party) |>
  summarize(mean_estimated_voting = mean(estimated_voting, na.rm = TRUE), .groups = "drop")|>
   mutate(anno = year(date_elec), mes = month(date_elec)) |> 
    select(anno, mes, party, mean_estimated_voting)

# Now we'll merge the two datasets (election results and survey estimates)
# Before, we change some party names so they fit
survey_estimates <- survey_estimates |> 
  mutate(party = case_when(
    party == "CIUDADANOS PARTIDO DE LA CIUDADANÍA" ~ "CIUDADANOS PARTIDO DE LA CIUDADANÍA",
    party == "PARTIDO POPULAR" ~ "PARTIDO POPULAR",
    party == "PARTIDO SOCIALISTA OBRERO ESPAÑOL" ~ "PARTIDO SOCIALISTA OBRERO ESPAÑOL",
    party == "UNIDAS PODEMOS-IU" ~ "PODEMOS-IU",
    party == "VOX" ~ "VOX",
        party == "EH-BILDU" ~ "EH BILDU",
    TRUE ~ party
  ))

#Merge
poll_error_data <- left_join(total_votes_by_party, survey_estimates, by = c("anno", "mes", "party"))

# Calculate the absolute error for each survey for each party
poll_error_data <- poll_error_data |> 
  mutate(error_percentage = abs(real_percentage - (mean_estimated_voting/100)))

# Print the data with the error (see the table below)
print(poll_error_data)

```

### Question 6

Which polling houses got it right the most and which ones deviated the most from the results? Check if works

```{r}
#| code-fold: true
library(dplyr)

#'final_survey_data' and 'final_election_data' are available from the previous steps

# 1. Calculate real vote percentages from election data
real_votes <- final_election_data |>
  group_by(anno, mes, party) |>
  summarize(total_ballots = sum(ballots, na.rm = TRUE), .groups = "drop") |>
    group_by(anno, mes) |>
      mutate(total_ballots_election = sum(total_ballots, na.rm = TRUE)) |>
  ungroup() |>
  mutate(real_percentage = total_ballots / total_ballots_election) |>
  select(anno, mes, party, real_percentage) |>
    mutate(mes = as.integer(mes)) # Convert mes to integer


# 2. Join real vote percentages with survey data
# Adjust party names in survey data to match election data before the join
final_survey_data_adjusted <- final_survey_data |>
    mutate(party = case_when(
        party == "CIUDADANOS PARTIDO DE LA CIUDADANÍA" ~ "CIUDADANOS PARTIDO DE LA CIUDADANÍA",
        party == "PARTIDO POPULAR" ~ "PARTIDO POPULAR",
        party == "PARTIDO SOCIALISTA OBRERO ESPAÑOL" ~ "PARTIDO SOCIALISTA OBRERO ESPAÑOL",
        party == "UNIDAS PODEMOS-IU" ~ "PODEMOS-IU",
        party == "VOX" ~ "VOX",
        party == "EH-BILDU" ~ "EH BILDU",
        TRUE ~ party
      )) |>
    mutate(mes = month(date_elec), anno= year(date_elec))


survey_with_real <- final_survey_data_adjusted |>
  left_join(real_votes, by = c("anno", "mes", "party"))

# 3. Calculate the error for each pollster
pollster_error <- survey_with_real |>
  group_by(pollster) |>
  summarize(average_error = mean(abs(estimated_voting - (real_percentage * 100)), na.rm = TRUE), .groups = "drop")

# 4. Identify the pollsters with the smallest average error (most accurate)
best_pollsters <- pollster_error |>
  arrange(average_error)

# 5. Identify the pollsters with the biggest average error (least accurate)
worst_pollsters <- pollster_error |>
  arrange(desc(average_error))

# 6. Print the results
print("Best Polling Houses (Most Accurate):")
print(head(best_pollsters))
```

Most Accurate Polling Houses were: OPINA, VOX PÚBLICA and METRA SEIS, with average errors of +/- 3-4% meaning that pollster's estimates were, on average, 3-4 percentage points different from the actual results. This error refers to a general average over the years we have considered in our data set (2008-2018)

```{r}
#| code-fold: true

print("Worst Polling Houses (Least Accurate):")
print(head(worst_pollsters))
```

And the worst were: MYWORD, HAMALGAMA MÉTRICA ans SW DEMOSCOPIA where the pollster's estimates were, on average, 7.5 percentage points different from the actual results.

### Original Question 1
Q Julia ponga la pregunta exacta pero es sobre PARTICIPACIÓN

```{r}
#| code-fold: true
# 1. Prepare Election Data
election_data_turnout <- final_election_data |>
  filter(tipo_eleccion == "02") |> #Filter for congressional elections
  mutate(election_year = year(make_date(anno, mes))) #Extract election year

# 2. Calculate Total Census and Votes per Year
turnout_by_year <- election_data_turnout |>
    group_by(election_year) |>
    summarise(
      total_census = sum(censo, na.rm = TRUE),  # Calculate total census
      total_votes = sum(votos_candidaturas, na.rm = TRUE) + sum(votos_blancos, na.rm = TRUE), #Calculate total valid votes
        .groups = "drop"
        )

# 3. Calculate Turnout Percentage
turnout_by_year <- turnout_by_year |>
    mutate(turnout_percentage = (total_votes / total_census) * 100) 

# 4. Display Results (just in case)
print("Voter Turnout by Year:")
print(turnout_by_year)
```

```{r}
#| code-fold: true
# 5. Visualize Results
ggplot(turnout_by_year, aes(x = election_year, y = turnout_percentage)) +
  geom_line(color = "darkblue", linewidth = 1.2) +
  geom_point(color = "darkblue", size = 3) +
    geom_text(aes(label = sprintf("%.1f%%", turnout_percentage), # Add text labels with percentage
                vjust = ifelse(election_year %in% c(2011, 2016), 1.5, -1)), # Adjust vjust for 2011 and 2016
               size = 3, color = "darkblue") + 
  scale_x_continuous(breaks = unique(turnout_by_year$election_year)) +
  scale_y_continuous(limits = c(65, 80), breaks = seq(65, 80, 5), expand = c(0,0)) +
    labs(title = "Turnout Percentage in Spanish Congressional Elections (2008-2019)",
         x = "Election Year",
         y = "Turnout Percentage") +
  theme_minimal() +
    theme(panel.grid.major.x = element_blank(),
          panel.grid.minor.x = element_blank(),
         panel.grid.major.y = element_line(color = "gray", linewidth = 0.3),
          panel.grid.minor.y = element_blank())
```

### Original Question 2

```{r}
#| code-fold: true

#regresión multinomial por cada partido importante: PSOE, podemos, PP, VOX añadiendo base de datos y viendo q variables influyen en el voto (esto lo hice en la carrera exactamente igual asi q no creo q tengamos problemas)

```


### Original Question 3


```{r}
votes_2019 <- votes_by_party_all |> 
  filter(anno == 2019, mes == 11)

national_total_2019 <- sum(votes_2019$total_votes)

votes_2019 <- votes_2019 |> 
  group_by(siglas) |> 
  summarise(votes_by_party_2019 = sum(total_votes)) |> 
  mutate(percentage_votes = (votes_by_party_2019 / national_total_2019) * 100)

votes_2019 <- votes_2019 |> rename(party = siglas)

votes_2019

```

```{r}
escanos_total <- 350

party <- c("BNG", "CIU", "CS", "EAJ-PNV", "EH-BILDU", "ERC", "MP", "PODEMOS-IU", "PP", "PSOE", "Other")

escanos_asigned <- c(1, 0, 10, 6, 5, 13, 3, 35, 89, 120, 81)

percentage_escanos <- (escanos_asigned / escanos_total) * 100

results_2019_nov <- data.frame(
  party = party,
  escanos_asigned = escanos_asigned,
  percentage_escanos = percentage_escanos
)

print(results_2019_nov)
```

```{r}
total_votes_2019 <- sum(votes_2019$votes_by_party_2019)

votes_2019$percentage_votes <- (votes_2019$votes_by_party_2019 / total_votes_2019) * 100

comparison <- merge(results_2019_nov, votes_2019[, c("party", "percentage_votes")], by = "party")

print(comparison)
```

```{r}
library(ggplot2)
library(forcats)
library(tidyr)

table_long <- pivot_longer(
  comparison,
  cols = c(percentage_escanos, percentage_votes),
  names_to = "type",
  values_to = "percentage"
)

table_long$party <- fct_reorder(table_long$party, table_long$percentage, .fun = max, .desc = TRUE)

ggplot(table_long, aes(x = party, y = percentage, fill = type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Comparison of percentage of escaños vs votes by party",
    x = "Party",
    y = "Percentage (%)",
    fill = "Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )
```

### Original Question $: algún tipo de loop con final_surveys
